---
title: "Births & conceptions in Belgium"
output: 
  flexdashboard::flex_dashboard:
    theme: lumen
    source: embed
    storyboard: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(flexdashboard)
library(dplyr)
library(ggplot2)
library(ggiraph)
library(plotly)

births_conceptions_be = readRDS('births_conceptions_be.rds')
theme_set(theme_minimal())

# helpers ----------------------------------------------------------

scale_date_custom = function() scale_x_date(
  date_breaks = '2 years', date_labels = "'%y", 
  limits = as.Date(c('1992-01-01', '2020-01-01'))
)

# knitr::opts_chunk$set(eval = FALSE)
```

### Births in Belgium between 1992 - 2020; or how you should **NOT** visualize the data.

```{r}
gg = ggplot(births_conceptions_be, 
            aes(x=date, y = nbirths))+
  geom_line()+
  geom_smooth()+
  scale_date_custom()+
  scale_y_continuous(limits = c(0, 500))+ 
  labs(
    x = 'Year', 
    y = 'Number of births', 
    title = 'Daily births in Belgium<br><sup>Between 1992 and 2020</sup>', 
  )

plotly::ggplotly(gg) %>% 
  plotly::style(traces = 2:3, hoverinfo = 'none')
```

------------------------------------------------------------------------

Recently, one of my colleagues shared an open source dataset on the
daily number of births in Belgium. Exploring the data, it quickly became
clear birthdays are less random than you might think.

In the next few blocks, I hope to have shared some insights. ---

Let's start with the time series plot on the left. We can notice that
there is 1. quite some **variation** in the number of births on any
given moment 2. an indication for a **change in time**

The plot isn't ideal though... you'll see why :-)

------------------------------------------------------------------------


### Our first improvement

```{r}
gg = ggplot(births_conceptions_be, 
            aes(x=date, y = nbirths))+
  geom_point(aes(weekday=weekday), size=.1)+
  geom_smooth(se = FALSE)+
  scale_date_custom()+
  scale_y_continuous(limits = c(0, 500))+ 
  scale_color_discrete()+
  labs(
    x = 'Year',
    y = 'Number of births', 
    color = 'Weekday', 
    title = 'Daily births in Belgium'
  )
plotly::ggplotly(gg) %>% 
  style(hoverinfo = 'none', traces = 2)
```

------------------------------------------------------------------------

Clearly there is something going on...

If you hover over the scatter cloud you quickly see an indication for
the reason of such a difference.

Or... just plain reasoning about impacting factors should work perfectly
as well in this simple case :)

### The effect of weekend on births

```{r}
gg = ggplot(births_conceptions_be, 
            aes(x=date, y = nbirths, color = weekday))+
  geom_point(size=.1)+
  geom_smooth(se = FALSE)+
  scale_date_custom()+
  scale_y_continuous(limits = c(0, 500))+ 
  scale_color_discrete()+
  labs(
    x = 'Year',
    y = 'Number of births', 
    color = 'Weekday', 
    title = 'Daily births in Belgium<br><sup>Presence of a weekend effect</sup>'
  )


plotly::ggplotly(gg) %>% 
  style(hoverinfo = 'none', traces =  8:14)
```

------------------------------------------------------------------------

We could expect that births were lower in the weekend. However, that
there is such a large difference might be surprising nonetheless.

------------------------------------------------------------------------

```{r, eval=FALSE, echo=FALSE}
births_conceptions_be %>% 
  filter(!is.na(nbirths)) %>% 
  group_by(weekend = weekday %in% c('Sat', 'Sun')) %>% 
  summarise(avg_daily_births = mean(nbirths))
```

Over the years, the average number of births during a weekend day is
215. For weekdays, it is 373. On average, one is **1.73 times more
likely to be born during a weekday**.\*

------------------------------------------------------------------------

\**Simplified example which does not take year, holidays etc. into
account.*

### An alternative representation: **birthday heatmap**

```{r}
hovertemplate  = "<b>{yy}</b>
<i>{weekday} {dd} {month.abb[mm]}</i>

No. births: <b>{nbirths}</b>
Birthday popularity: <b>{round(bd_pop, 2)}</b>{tooltip_holiday}"

heatmapdata = births_conceptions_be %>% 
  filter(!is.na(nbirths)) %>% 
  group_by(yy) %>% 
  mutate(
    bd_pop = nbirths / mean(nbirths), 
    tooltip_holiday = ifelse(is.na(holiday), '', paste0('\nHoliday: <b>', holiday, '</b>')),
    tooltip = glue::glue(hovertemplate),
    Year = yy
  )

gg = ggplot(heatmapdata %>% filter(yy == 1992), 
       aes(fill=bd_pop, x = dd, y = mm, frame = Year, text=tooltip))+
  geom_tile()+
  scale_fill_gradient2(midpoint = 1, breaks = seq(.4, 1.4, .2))+
  scale_y_discrete(limits = rev)+
  labs(
    title = "Birthday popularity in 1992",
    x = 'Day', y = "Month", 
    fill = 'Birthday\npopularity')+
  theme(
    panel.background = element_blank(), 
    panel.grid = element_blank(),
    axis.ticks = element_blank()
  )


plotly::ggplotly(gg, tooltip = 'text') %>% 
  layout(
    hoverlabel=list(bgcolor="white")
  )
```

------------------------------------------------------------------------

So, we've seen a time series and a scatter plot. The same data can be
presented by means of the heatmap on the left. Purple days indicate more
births on those days compared to average in that year. The opposite
holds for the red tiles.

------------------------------------------------------------------------

Again, we can identify the weekend effect on this plot. The red tiles
are mostly concentrated with two less common birthdays next to each
other.

### The heatmap, in motion

```{r}


gg = ggplot(heatmapdata, 
       aes(fill=bd_pop, x = dd, y = mm, frame = Year, text=tooltip))+
  geom_tile()+
  scale_fill_gradient2(midpoint = 1, breaks = seq(.4, 1.4, .2))+
  scale_x_discrete(position='top')+
  scale_y_discrete(limits = rev)+
  labs(x = 'Day', y = "Month", 
       fill = 'Birthday\npopularity')+
  theme(
    panel.background = element_blank(), 
    panel.grid = element_blank(),
    axis.ticks = element_blank()
  ) 

plotly::ggplotly(gg, tooltip = 'text') %>% 
  plotly::animation_slider(
    pad = list(t = 10)
  ) %>%
  layout(
    hoverlabel=list(bgcolor="white"),
    xaxis=list(side='top')
  ) %>%
  animation_button(
    visible = TRUE,
    pad = list(r = 20)
  )

```

------------------------------------------------------------------------

Similar to the previous visual, we now have a sequence of heatmaps for
consecutive years. Playing the animation will also make clear that the
red blocks just shift horizontally as weekdays shift each year.

------------------------------------------------------------------------

Hovering over the tiles will give you some details. If you're born in
1992 or later, you can check how common your birthday is during that
year.

------------------------------------------------------------------------

In one of the next blocks, this visualization will help us to avoid
making false conclusions!

------------------------------------------------------------------------

GIF available [here](file.gif)

### Patterns in births = patterns in conceptions. *When do people conceive babies?*  {data-commentary-width="500"}


```{r}
# make data
spiral_data =
  births_conceptions_be %>% 
  filter(!is.na(nconceptions)) %>% 
  filter(!(mm == 2 & dd == 29)) %>%
  group_by(yy, mm) %>% 

  # daily avg per yymm
  mutate(
    yymm_mean = round(mean(nconceptions))
  ) %>% 
  group_by(yy) %>% 
  # comparison yymm daily average to yy daily average and create rank
  mutate(
    month_conc_popularity = 13 - dense_rank(yymm_mean),
    tooltip = glue::glue("<b>{month.name[mm]} {yy}</b>\n\nAvg daily conceptions: {yymm_mean}\nConception rank within year: <b>{month_conc_popularity}</b>"),
  ) %>% 
  ungroup() %>% 
  
  # helper variable for polar spiral
  mutate(
    day_of_year_cumsum = cumsum(day_of_year)
  )

# tickpos years; angular axis
tickvals_years = spiral_data %>% 
  filter(dd == 1, mm == 1, yy %in% c(1995, 2020)) %>% 
  pull(day_of_year_cumsum)


# tickpos months
tickvals_months = as.integer(seq(as.Date('1970-01-15'), as.Date('1970-12-15'), by = 'month'))

plot_ly(
    spiral_data, 
    colors = colorRamp(c('darkblue', 'white', 'darkred')),
    type = 'scatterpolar',
    mode = 'markers',
    theta = ~day_of_year, 
    r = ~day_of_year_cumsum, 
    text = ~tooltip, 
    color = ~factor(month_conc_popularity),
    hoverinfo = 'text',
    marker = list(size = 4)
  ) %>%
  layout(
    legend = list(title = list(text = 'Daily average\nof conceptions\n<sup>relative to month within year\n\n1: most fertile\n12: least fertile</sup>\n')),
    polar = list(
      angularaxis  = list(
        showline = FALSE,
        showgrid = FALSE,
        rotation = 90,
        direction = "clockwise",
        tickmode="array",
        tickvals = tickvals_months,
        ticktext = month.abb,
        ticks = ''
      ),
      radialaxis = list(
        # angle = 90,
        # tickangle = 90, 
        showline = FALSE,
        showgrid = FALSE,
        ticks = '', 
        tickmode = 'array', 
        tickvals = tickvals_years, 
        ticktext = c("<b>1995</b>", "<b>2020</b>"), 
        tickfont = list(color = "black")
      )
    ),
    xaxis = list(
    'showgrid' = FALSE, # thin lines in the background
    'zeroline' = FALSE, # thick line at x=0
    'visible' = FALSE   # numbers below
    ),
    yaxis = list(
    'showgrid' = FALSE, # thin lines in the background
    'zeroline' = FALSE, # thick line at x=0
    'visible' = FALSE   # numbers below
    )
  )

```

------------------------------------------------------------------------

<!-- TODO -->
Last but not least, I created a spiral plot.

For each month, the color indicates




------------------------------------------------------------------------

\**The choice of a 'spiral plot' for the conceptions was
non-intentional. This visualization was somewhat inspired by this
visualization on
[COVID](https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html)*

------------------------------------------------------------------------


Details on the data

The dataset is made available Statbel, the Belgian statistical office.
You can access it
[here](https://statbel.fgov.be/en/open-data/number-births-day).

The number of births per day between 1 Jan 1992 and 31 Dec 2020 are
reported.

Based on the dates, some extra features are included such as the
weekday, day of year and Belgian historical holidays that were scraped
[here](https://www.kalender-365.be/feestdagen/1991.html)